A conditional permutation-based approach to test confounder and center effects in machine learning models

### 2021-10-21 manuscript.tex

\title{A conditional permutation-based approach to test confounder effect and center-bias in machine learning models}

@@

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confounder-bias significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal and display a remarkable robustness and a high statistical power in simulations.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset provided statistical evidence for several models being significantly confounded. Some of these confounder-biases were previously unreported and found to be hard to fully mitigate with state-of-the-art confound mitigation approaches.
The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}

@@

Predictive modelling and supervised learning methods have recently became increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Generalizability across contexts and populations and biomedical validity (i.e. whether the model is driven by biomedically relevant signal) are also crucial requirements for candidate biomarkers \citep{woo2017building}.

Variables that can induce spurious associations between the predictor variables (features) and the prediction target - often referred to as confounders - can be detrimental to the model's biomedical validity and generalizability. Many types of confounders can be distinguished.
First, measurement artifacts can obviously be considered as confounders. For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}. Similarly, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. For instance, resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}. As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}. 
Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target. Such models will display worse-than-expected performance when applied on data from new centers/batches.

Various data harmonization approaches have been proposed to mitigate confounder-bias, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}. However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful and robust statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches. Such methods, however, have to tackle the fact that the outputs of predictive models often follow non-normal distributions (due to e.g. regularization) \citep{garcia2009study, kristensen2017whole}.

>>>>

\title{A conditional permutation-based approach to test confounder and center effects in machine learning models}

@@

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confounder effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded model}, respectively.
The proposed methods provide a strict control for Type-I errors and high statistical power, even in the presence of typical non-linear and non-normal dependencies between the target variable and the model output.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confoudner effects that were previously unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.
The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}

@@

Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Biomedical validity (i.e. whether the model is driven by biomedically relevant signal) and generalizability across contexts and populations are also crucial requirements for candidate biomarkers \citep{woo2017building}.

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - induced by so-called confounder effects - can be detrimental to the model's biomedical validity and generalizability. Many types of confounder effects can be distinguished.

First, measurement artifacts can obviously be considered as confounders. For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}. Likewise, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.

Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}. As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}.

Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target. Such models will display worse-than-expected performance when applied on data from new centers/batches.

Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}. However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful and robust statistical tests to quantify the confounding effect of certain variables in predictive models could substantially foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches. Such methods, however, have to tackle the fact that the conditional distribution of the models' output, given the target, is often non-normal and non-linear, as a consequence of e.g. model regularization \citep{garcia2009study, kristensen2017whole}.
